---
title: "Module 8 Project Report"
author: "FeatheR"
date: "2023-11-17"
output: html_document
bibliography: BIOL3140.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rgbif)
library(tidyverse)
library(MuMIn)
library(rnoaa)
library(data.table)
library(ggmap)
library(usmap)
library(sf)
library(magick)
library(cowplot)
library(lme4)
library(car)
library(data.table)
```

##Introduction:
The aim of this project is to use eBird and meteorological data to study the effect of local weather conditions on the arrival time of trans-Gulf migrant (TGM) species in Massachusetts. Specifically, we are examining if arrival time of five different species varies according to temperature and wind variables along migration route for these TGMs migrating to Massachusetts. Once considering that arrival time may vary with meteorological conditions, we will also consider what role climate change will potentially play in the population status of TGMs arriving in MA during the spring. Many researchers have studied the relationship between weather and migration patterns, including [@richardson1978timing], who concluded that maximum numbers migrate with fair weather, with tailwinds and with temperature, pressure and humidity conditions that accompany tailwinds. 

##Methods:
We began by downloading eBird’s species occurrence data from the Global Biodiversity Information Facility (GBIF) and selecting 5 TGMs to analyse. We then compiled weather using rnoa which contains functions that interact with NOAA’s National Climatic Data Center’s API. Fot analysis, we computed arrival time for each species by finding the day that corresponds to when 25% of all the individuals have arrived in a given year and modeled this process as a logistics curve. We then found the Julian day each year that is closest to 0.25 of the population arriving and plotted how this arrival day varied with year. When preparing our weather data, we scaled the wind direction to 180 degrees and essentially selected the weather columns we want to work with. With two joined tibbles each containing eBird and weather data, we were finally able to use linear mixed-effect modeling to evaluate the effect of weather parameters on arrival time. The Anova() functiong was key in assessing the significance of the fixed effects we’ve used and the Dredge() function allowed us to determine our best-fit nested model.

##Results:

```{r querying GBIFs API, echo=FALSE, cache = TRUE}
species <- c("Piranga olivacea", "Leiothlypis ruficapilla", "Dolichonyx oryzivorus", "Megaceryle alcyon","Chordeiles minor")
year <- paste0("1990",",","2022")
month <- paste0("3",",","6")

data.l <- list() #making empty list to store data

for (s in species) {
  obs.num <- occ_data(scientificName = s, year = year, month = month, limit = 0, country = "US", basisOfRecord = "HUMAN_OBSERVATION", stateProvince = "Massachusetts")$meta$count
  data.l[[paste0(s)]] <- occ_data(scientificName = s, year = year, month = month,
                               limit = obs.num, country = "US",
                               basisOfRecord = "HUMAN_OBSERVATION",
                               stateProvince = "Massachusetts")[[2]]
}

dat <- rbindlist(data.l,fill=T)

head(dat)

saveRDS(dat, "massbird.data.RDS")

dat%>%
  group_by(year,species)%>%
  summarise(count=sum(individualCount,na.rm = T))%>%
  ggplot(aes(x=year,y=count,col=species))+geom_point()
```

``` {r querying NOAAs NCDC API, echo = FALSE}
options(noaakey = "pXwWlsmmheEQSwUlHjihgmoZKLOCKhvE")

sts <- c(
  "GHCND:USW00013894", #Mobible, AL 2k away about 10 days away @200 km/day
  "GHCND:USW00013881", #Charlotte, NC 1000 km away about 6 days away @200 km/day
  "GHCND:USW00014739") #Boston

bos <- ncdc_stations(stationid = "GHCND:USW00014739")

sta.data <- bind_rows(
  lapply(sts,function(x) ncdc_stations(stationid = x)$data)) %>%
  mutate(usmap_transform(.,input_names = c("longitude","latitude"), output_names = c("longitude.1", "latitude.1"))) %>% #join transformation of lat/long for projection with usmap
  mutate(name = str_sub(name, -5,-4)) %>% #simplify the name column, grab just the state
  mutate(migr.day = c(10,5,0)) %>% #so we can look at wind speed 0, 5 or 10 days before arrive in boston
  separate(id,into = c("station.type","id")) #need to cut station type out from station id number

print(sta.data)

plot_usmap(include = c(.northeast_region, .south_region, .east_north_central)) + geom_point(data = sta.data, aes(x = longitude.1, y = latitude.1, col = name), size = 5) + geom_label(data = sta.data, aes(x = longitude.1, y=latitude.1, col = name, label = name), size = 5, nudge_x = 1e6*0.25) + theme(legend.position = "none")

#Now we want weather data from these stations during the spring of the years we have good data for (pretty much 2000 on)
weather.data <- meteo_pull_monitors(sta.data$id,date_min = "2000-01-01")
#tmin = minimum temperature in tenths of ºC
#tmax = maximum temperature in tenths of ºC
#awnd = average wind velocity in m^(-s)
#wdf2 = direction of fastest 2-minute wind
```

```{r preparing eBird data}
po <- dat %>%
  filter(species == "Piranga olivacea") %>%
  group_by(year) %>%
  mutate(date = as.Date(paste0(year,"-",month,"-",day)),
         j.day = julian(date, origin = as.Date(paste0(unique(year),"-01-01")))) %>%
  group_by(species, year, j.day, date) %>%
  summarise(day.tot = sum(individualCount, na.rm=T)) %>%
  group_by(species, year) %>%
  mutate(prop = cumsum(day.tot/sum(day.tot,na.rm = T))) %>%
  filter(year>1999)

lr <- dat %>%
  filter(species == "Leiothlypis ruficapilla") %>%
  group_by(year) %>%
  mutate(date = as.Date(paste0(year,"-",month,"-",day)),
         j.day = julian(date, origin = as.Date(paste0(unique(year),"-01-01")))) %>%
  group_by(species, year, j.day, date) %>%
  summarise(day.tot = sum(individualCount, na.rm=T)) %>%
  group_by(species, year) %>%
  mutate(prop = cumsum(day.tot/sum(day.tot,na.rm = T))) %>%
  filter(year>1999)

do <- dat %>%
  filter(species == "Dolichonyx oryzivorus") %>%
  group_by(year) %>%
  mutate(date = as.Date(paste0(year,"-",month,"-",day)),
         j.day = julian(date, origin = as.Date(paste0(unique(year),"-01-01")))) %>%
  group_by(species, year, j.day, date) %>%
  summarise(day.tot = sum(individualCount, na.rm=T)) %>%
  group_by(species, year) %>%
  mutate(prop = cumsum(day.tot/sum(day.tot,na.rm = T))) %>%
  filter(year>1999)

ma <- dat %>%
  filter(species == "Megaceryle alcyon") %>%
  group_by(year) %>%
  mutate(date = as.Date(paste0(year,"-",month,"-",day)),
         j.day = julian(date, origin = as.Date(paste0(unique(year),"-01-01")))) %>%
  group_by(species, year, j.day, date) %>%
  summarise(day.tot = sum(individualCount, na.rm=T)) %>%
  group_by(species, year) %>%
  mutate(prop = cumsum(day.tot/sum(day.tot,na.rm = T))) %>%
  filter(year>1999)

cm <- dat %>%
  filter(species == "Chordeiles minor") %>%
  group_by(year) %>%
  mutate(date = as.Date(paste0(year,"-",month,"-",day)),
         j.day = julian(date, origin = as.Date(paste0(unique(year),"-01-01")))) %>%
  group_by(species, year, j.day, date) %>%
  summarise(day.tot = sum(individualCount, na.rm=T)) %>%
  group_by(species, year) %>%
  mutate(prop = cumsum(day.tot/sum(day.tot,na.rm = T))) %>%
  filter(year>1999)

#arrival of these migrants follows a logistic process when we consider the proportion of the population that has arrived!

#lets use logistic modeling:
po.pred <- po %>%
  group_by(year) %>%
 summarize(pred = predict(nls(prop ~ SSlogis(j.day, Asym, xmid, scal)), newdata = data.frame(j.day = min(j.day):max(j.day))), #predict the logistic curve for each species
   j.day=min(j.day):max(j.day),) %>%
  left_join(po %>% dplyr::select(j.day,date)) #add date back to tibble

po %>%
  ggplot(aes(j.day,prop)) + geom_point(aes=0.3) + geom_line(data = po.pred, aes(x = j.day, y = pred), col="blue", size=2) + facet_wrap(year~.)

po.arrive.date <-po.pred%>%
  group_by(year) %>%
  filter(j.day==j.day[which.min(abs(pred-0.25))])

po.arrive.date %>%
  ggplot(aes(year,j.day))+ geom_point()
```

```{r preparing weather data}
weather.data <- weather.data %>%
  mutate(year = as.integer(str_sub(date, 1, 4)), dat = as.Date(date)) %>%
  group_by(year) %>%
 mutate(j.day = julian(date, origin = as.Date(paste0(unique(year), "-01-01"))), 
  date2 = date,
  wdir.rad = (180-abs(wdf2-180))*pi/180,
  wvec = cos(wdir.rad)*-1*awnd) %>% 
  dplyr::select(id, year, date2, j.day, tmin, tmax, wvec) %>% 
  left_join(sta.data %>% select(id, name, migr.day)) %>% 
  mutate(j.day = j.day + migr.day)

po.arr.weath <- po.arrive.date%>%
  left_join(weather.data)%>%
  left_join(po %>% dplyr::select(year,date,j.day))

weather.wk <-weather.data %>% 
  group_by(year, name) %>% 
  mutate(wk.tmin = frollmean(tmin, n=14, align="right"),
         wk.tmax = frollmean(tmax, n=14, align="right"),
         wk.wvec = frollmean(wvec, n=14, align="right"))%>%
  dplyr::select(j.day,date2,name,wk.tmin,wk.tmax,wk.wvec)

po.arr.weath2 <- po.arrive.date %>%
  left_join(weather.wk)

po.arr.weath2
```

```{r linear mixed effect models}

#Linear mixed-effect modeling of arrival day as it varies with these weather variables for both weather data sets (single-day and two-week average)
#Model-testing operations of both data sets using dredge().
#Anova testing of the best-fit models from both data sets.

#weather at 0, 5, and 10 days away from arrival
po.lmer <- lmer(j.day~tmin*tmax*wvec+(1|name),po.arr.weath,na.action = "na.fail")
Anova(po.lmer)

#0Mean two week weather preceding arrival
po.lmer2 <- lmer(j.day~wk.tmin*wk.tmax*wk.wvec+(1|name),po.arr.weath2,na.action = "na.fail")
Anova(po.lmer2)

po.arr.aic <- dredge(po.lmer2,fixed = c("wk.tmin","wk.tmax","wk.wvec"),)

po.kb <- kable(po.arr.aic[1:4,],caption = "Fit values for nested models of the most complicated lme model")

kable_styling(po.kb)

best.lmer <-  lmer(j.day~wk.tmin+wk.tmax+wk.wvec+(1|name),po.arr.weath2,na.action = "na.fail")

Anova(best.lmer)
```

##Discussion:
How does your analysis contribute to, challenge, or refine previous hypothesis concerning the role that climatic variables play in long-distance migration in passerine birds?

According to the linear mixed-effects models, the arrival time of TGMs in MA may indeed vary with meteorological conditions along their migration route. The model investigates the effect of all three variables (tmax, tmin, wvec) in relation to arrival time. The p-values for tmax ((0.0634) and tmin (0.6455) in the daily weather model indicate that maximum temperature has a near-significant effect on arrival time (p < 0.1), suggesting a possible trend, while minimum temperature does not have a statistically significant effect. The wind variable (wvec) also does not show a significant effect (p = 0.7007).

This analysis provides evidence that may suggest a role for maximum temperature in influencing migration timing. Thus, climate change will most likely affect the arrival date of migrartory birds. As global warming progresses, the maximum temperature will increase as well, which may eventually become significant to the arrival time.

In [@knudsen2011challenging]'s paper on birds' spring migration in response to climate change, it was found that birds advance their spring migration in response to climate change and climate change affects migration distance and routes. Our data support this claim in which a max temperature may contribute to the arrival time. 

##Author Contributions: 
The group communicated outside of class to split the work up. Jaan inserted the necessary code chunks. Elisabeth created the bibliography and crafted the Introduction, Methods, and Author Contributions. Yingshan finished the linear mixed effect coding and wrote subsequent interpretation for the discussion.

##References:
